---
title: "Create Observational and Interventional experiments on Alarm Network"
author: "Megha"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this file, we simulate obseravtional and interventional experiment from the synthetic network alarm from the bnlearn package. We conduct two experiment one generating alarm_mut1 with one intervention amd alarm_mut2 with two intervention. This is done in the same way to replicate the process of Sachs data generation. We also add unknown "fat-hand" intervention by changing values of one or more of the target node's children.

```{r include=FALSE, cache=FALSE}
#loading the packages
install.packages("bnlearn")
library(bnlearn)
source("http://bioconductor.org/biocLite.R")
biocLite("Rgraphviz")
library(Rgraphviz)
library(ggplot2)
source("https://bioconductor.org/biocLite.R")  ##Use to load gRain package since RBGL is removed from CRAN repository
biocLite("RBGL")
library(gRain) 
library(gRbase)
```

```{r}
# load the alarm dataset.
data(alarm)
# create and plot the network structure.
dag_alarm = model2network(paste0("[HIST|LVF][CVP|LVV][PCWP|LVV][HYP][LVV|HYP:LVF][LVF]",
  "[STKV|HYP:LVF][ERLO][HRBP|ERLO:HR][HREK|ERCA:HR][ERCA][HRSA|ERCA:HR][ANES]",
  "[APL][TPR|APL][ECO2|ACO2:VLNG][KINK][MINV|INT:VLNG][FIO2][PVS|FIO2:VALV]",
  "[SAO2|PVS:SHNT][PAP|PMB][PMB][SHNT|INT:PMB][INT][PRSS|INT:KINK:VTUB][DISC]",
  "[MVS][VMCH|MVS][VTUB|DISC:VMCH][VLNG|INT:KINK:VTUB][VALV|INT:VLNG]",
  "[ACO2|VALV][CCHL|ACO2:ANES:SAO2:TPR][HR|CCHL][CO|HR:STKV][BP|CO:TPR]"))
```

```{r}
#Sampling some observational data points from alarm dataset
alarm_mut1_obs1 = alarm[sample(nrow(alarm), 2000), ]
alarm_mut1_obs2 = alarm[sample(nrow(alarm), 2000), ]
alarm_mut1_obs3 = alarm[sample(nrow(alarm), 2000), ]


#Adding the INT column
alarm_mut1_obs1 = cbind(alarm_mut1_obs1, INT = 0)
alarm_mut1_obs2 = cbind(alarm_mut1_obs2, INT = 0)
alarm_mut1_obs3 = cbind(alarm_mut1_obs3, INT = 0)
```

```{r}
#First we start by learning parameters of the Bayesian Network given the DAG
fitted = bn.fit(dag_alarm, alarm, method = "bayes")
```

Performing queries using gRain package on a mutilated Bayesian Network where we performed a perfect intervention by setting PMB as "TRUE". This is similar to a lab experiment where we intervene PMB like a knock-out experiment in the real world.

```{r}
#Adding perfect intervening at node "PMB" and fat-hand intervention at node "SHNT"
alarm_mut1_network = mutilated(fitted, evidence = list(PMB = "TRUE", SHNT= "HIGH"))

#Sampling experimental data
alarm_mut1_exp1 = rbn(alarm_mut1_network,2000)


#Adding the INT column
alarm_mut1_exp1 =  cbind(alarm_mut1_exp1, INT = 21)

#---------------------------------------------------------------------------------------

#Adding perfect intervening at node "VTUB"
alarm_mut1_network = mutilated(fitted, evidence = list(VTUB = "LOW"))

#Sampling experimental data
alarm_mut1_exp2 = rbn(alarm_mut1_network,2000)

#Adding the INT column
alarm_mut1_exp2 = cbind(alarm_mut1_exp2 , INT = 36)

#---------------------------------------------------------------------------------------

#Adding perfect intervening at node "VALV" and fat-hand intervention at node "ACO2"
alarm_mut1_network = mutilated(fitted, evidence = list(VALV = "LOW", ACO2= "NORMAL"))

#Sampling experimental data
alarm_mut1_exp3 = rbn(alarm_mut1_network,2000)

#Adding the INT column
alarm_mut1_exp3= cbind(alarm_mut1_exp3 , INT = 34)

#---------------------------------------------------------------------------------------

#Adding perfect intervening at node "ANES" 
alarm_mut1_network = mutilated(fitted, evidence = list(ANES = "TRUE"))

#Sampling experimental data
alarm_mut1_exp4 = rbn(alarm_mut1_network,2000)


#Adding the INT column
alarm_mut1_exp4= cbind(alarm_mut1_exp4 , INT = 20)

#---------------------------------------------------------------------------------------
#Adding perfect intervening at node "ERCA" 
alarm_mut1_network = mutilated(fitted, evidence = list(ERCA = "TRUE"))

#Sampling experimental data
alarm_mut1_exp5 = rbn(alarm_mut1_network,2000)


#Adding the INT column
alarm_mut1_exp5 = cbind(alarm_mut1_exp5 , INT = 30)

#---------------------------------------------------------------------------------------

#Adding perfect intervening at node "CCHL" and fat-hand intervention at node "HR"
alarm_mut1_network = mutilated(fitted, evidence = list(CCHL = "NORMAL", HR= "NORMAL"))

#Sampling experimental data
alarm_mut1_exp6 = rbn(alarm_mut1_network,2000)

#Adding the INT column
alarm_mut1_exp6 = cbind(alarm_mut1_exp6 , INT = 27)

```

```{r}
#Combining observation and experiment into one dataframe
alarm_mut1 = rbind(alarm_mut1_obs1, alarm_mut1_obs2, alarm_mut1_obs3, alarm_mut1_exp1, alarm_mut1_exp2, alarm_mut1_exp3, alarm_mut1_exp4, alarm_mut1_exp5, alarm_mut1_exp6)
write.csv(alarm_mut1,"data/Alarm data/alarm_mut1.csv",row.names=FALSE)
```
