---
title: "Create Observational and Interventional experiments on insurance Network"
author: "Megha"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this file, we simulate obseravtional and interventional experiment from the synthetic network insurance from the bnlearn package. We conduct two experiment one generating insurance_mut1 with one intervention amd insurance_mut2 with two intervention. This is done in the same way to replicate the process of Sachs data generation. We also add unknown "fat-hand" intervention by changing values of one or more of the target node's children.

```{r include=FALSE, cache=FALSE}
#loading the packages
install.packages("bnlearn")
library(bnlearn)
source("http://bioconductor.org/biocLite.R")
biocLite("Rgraphviz")
library(Rgraphviz)
library(ggplot2)
source("https://bioconductor.org/biocLite.R")  ##Use to load gRain package since RBGL is removed from CRAN repository
biocLite("RBGL")
library(gRain) 
library(gRbase)
```

```{r}
# load the insurance dataset.
data(insurance)
# create and plot the network structure.
dag_insurance = model2network(paste0("[Age][Mileage][SocioEcon|Age][GoodStudent|Age:SocioEcon]",
  "[RiskAversion|Age:SocioEcon][OtherCar|SocioEcon][VehicleYear|SocioEcon:RiskAversion]",
  "[MakeModel|SocioEcon:RiskAversion][SeniorTrain|Age:RiskAversion]",
  "[HomeBase|SocioEcon:RiskAversion][AntiTheft|SocioEcon:RiskAversion]",
  "[RuggedAuto|VehicleYear:MakeModel][Antilock|VehicleYear:MakeModel]",
  "[DrivingSkill|Age:SeniorTrain][CarValue|VehicleYear:MakeModel:Mileage]",
  "[Airbag|VehicleYear:MakeModel][DrivQuality|RiskAversion:DrivingSkill]",
  "[Theft|CarValue:HomeBase:AntiTheft][Cushioning|RuggedAuto:Airbag]",
  "[DrivHist|RiskAversion:DrivingSkill][Accident|DrivQuality:Mileage:Antilock]",
  "[ThisCarDam|RuggedAuto:Accident][OtherCarCost|RuggedAuto:Accident]",
  "[MedCost|Age:Accident:Cushioning][ILiCost|Accident]",
  "[ThisCarCost|ThisCarDam:Theft:CarValue][PropCost|ThisCarCost:OtherCarCost]"))
```

```{r}
#Sampling some observational data points from insurance dataset
insurance_mut1_obs = insurance[sample(nrow(insurance), 1000), ]

#intitializing additional column for Intervened target
Int_col = 0

#Adding the INT column
insurance_mut1_obs$INT = Int_col
```

```{r}
#First we start by learning parameters of the Bayesian Network given the DAG
fitted = bn.fit(dag_insurance, insurance, method = "bayes")
```

Performing queries using gRain package on a mutilated Bayesian Network where we performed a perfect intervention by setting Age as "Adult". This is similar to a lab experiment where we intervene L like a knock-out experiment in the real world.

```{r}
#Adding perfect intervening at node "Age" and fat-hand intervention at node "SocioEcon"
insurance_mut1_network = mutilated(fitted, evidence = list(Age = "Adult", SocioEcon = "Middle"))
jmutilated = compile(as.grain(insurance_mut1_network))

#Sampling experimental data
insurance_mut1_exp1 = rbn(insurance_mut1_network,1000)

#Intitializing additional column for the known intervened target
Int_col = 2

#Adding the INT column
insurance_mut1_exp1$INT = Int_col
```
```{r}
#Adding perfect intervening at node "DrivingSkill" and fat-hand intervention at node "DrivQuality"
insurance_mut1_network = mutilated(fitted, evidence = list(DrivingSkill = "SubStandard", DrivQuality = "Poor"))
jmutilated = compile(as.grain(insurance_mut1_network))

#Sampling experimental data
insurance_mut1_exp2 = rbn(insurance_mut1_network,1000)

#Intitializing additional column for the known intervened target
Int_col = 13

#Adding the INT column
insurance_mut1_exp2$INT = Int_col
```
```{r}
#Adding perfect intervening at node "MakeModel" 
insurance_mut1_network = mutilated(fitted, evidence = list(MakeModel = "Economy"))
jmutilated = compile(as.grain(insurance_mut1_network))

#Sampling experimental data
insurance_mut1_exp3 = rbn(insurance_mut1_network,1000)

#Intitializing additional column for the known intervened target
Int_col = 9

#Adding the INT column
insurance_mut1_exp3$INT = Int_col
```
```{r}
#Adding perfect intervening at node "Accident" 
insurance_mut1_network = mutilated(fitted, evidence = list(Accident = "Mild"))
jmutilated = compile(as.grain(insurance_mut1_network))

#Sampling experimental data
insurance_mut1_exp4 = rbn(insurance_mut1_network,1000)

#Intitializing additional column for the known intervened target
Int_col = 8

#Adding the INT column
insurance_mut1_exp4$INT = Int_col
```
```{r}
#Adding perfect intervening at node "OtherCarCost" 
insurance_mut1_network = mutilated(fitted, evidence = list(OtherCarCost = "Thousand"))
jmutilated = compile(as.grain(insurance_mut1_network))

#Sampling experimental data
insurance_mut1_exp5 = rbn(insurance_mut1_network,1000)

#Intitializing additional column for the known intervened target
Int_col = 21

#Adding the INT column
insurance_mut1_exp5$INT = Int_col
```


```{r}
#Combining observation and experiment into one dataframe
insurance_mut1 = rbind(insurance_mut1_obs, insurance_mut1_exp1, insurance_mut1_exp2, insurance_mut1_exp3, insurance_mut1_exp4, insurance_mut1_exp5)
write.csv(insurance_mut1,"data/insurance data/insurance_mut1.csv",row.names=FALSE)
```


Simulating data for with one observational and two interventinal experiments.
```{r}
#Sampling some observational data points from insurance dataset
insurance_mut2_obs1 = insurance[sample(nrow(insurance), 1000), ]
insurance_mut2_obs2 = insurance[sample(nrow(insurance), 1000), ]
insurance_mut2_obs3 = insurance[sample(nrow(insurance), 1000), ]

#intitializing additional column for Intervened target
Int_col = 0

#Adding the INT column
insurance_mut2_obs1$INT = Int_col 
insurance_mut2_obs2$INT = Int_col 
insurance_mut2_obs3$INT = Int_col 
```

Performing queries using gRain package on a mutilated Bayesian Network where we performed a perfect intervention by setting Lung cancer as "Yes" and Tuberculosis as "Yes". 

```{r}
#Adding perfect intervening at node "Age" and fat-hand intervention at node "SocioEcon"
insurance_mut2_network = mutilated(fitted, evidence = list(Age = "Adult", SocioEcon = "Middle"))
jmutilated = compile(as.grain(insurance_mut2_network))

#Sampling experimental data
insurance_mut2_exp1 = rbn(insurance_mut2_network,1000)

#Intitializing additional column for the known intervened target
Int_col = 2

#Adding the INT column
insurance_mut2_exp1$INT = Int_col

```
```{r}
#Adding perfect intervening at node "DrivingSkill" and fat-hand intervention at node "DrivQuality"
insurance_mut2_network = mutilated(fitted, evidence = list(DrivingSkill = "SubStandard", DrivQuality = "Poor"))
jmutilated = compile(as.grain(insurance_mut2_network))

#Sampling experimental data
insurance_mut2_exp2 = rbn(insurance_mut2_network,1000)

#Intitializing additional column for the known intervened target
Int_col = 13

#Adding the INT column
insurance_mut2_exp2$INT = Int_col
```
```{r}
#Adding perfect intervening at node "MakeModel" 
insurance_mut2_network = mutilated(fitted, evidence = list(MakeModel = "Economy"))
jmutilated = compile(as.grain(insurance_mut2_network))

#Sampling experimental data
insurance_mut2_exp3 = rbn(insurance_mut2_network,1000)

#Intitializing additional column for the known intervened target
Int_col = 9

#Adding the INT column
insurance_mut2_exp3$INT = Int_col
```
```{r}
#Adding perfect intervening at node "Mileage" and fat-hand intervention at "CarValue"
insurance_mut2_network = mutilated(fitted, evidence = list(Mileage = "Domino", CarValue = "TenThou"))
jmutilated = compile(as.grain(insurance_mut2_network))

#Sampling experimental data
insurance_mut2_exp4 = rbn(insurance_mut2_network,1000)

#Intitializing additional column for the known intervened target
Int_col = 11

#Adding the INT column
insurance_mut2_exp4$INT = Int_col
```
```{r}
#Adding perfect intervening at node "AntiTheft" 
insurance_mut2_network = mutilated(fitted, evidence = list(AntiTheft = "True"))
jmutilated = compile(as.grain(insurance_mut2_network))

#Sampling experimental data
insurance_mut2_exp5 = rbn(insurance_mut2_network,1000)

#Intitializing additional column for the known intervened target
Int_col = 19

#Adding the INT column
insurance_mut2_exp5$INT = Int_col
```
```{r}
#Adding perfect intervening at node "Airbag" 
insurance_mut2_network = mutilated(fitted, evidence = list(Airbag = "True"))
jmutilated = compile(as.grain(insurance_mut2_network))

#Sampling experimental data
insurance_mut2_exp6 = rbn(insurance_mut2_network,1000)

#Intitializing additional column for the known intervened target
Int_col = 25

#Adding the INT column
insurance_mut2_exp6$INT = Int_col
```
```{r}
#Adding perfect intervening at node "RiskAversion" 
insurance_mut2_network = mutilated(fitted, evidence = list(RiskAversion = "Normal"))
jmutilated = compile(as.grain(insurance_mut2_network))

#Sampling experimental data
insurance_mut2_exp7 = rbn(insurance_mut2_network,1000)

#Intitializing additional column for the known intervened target
Int_col = 4

#Adding the INT column
insurance_mut2_exp7$INT = Int_col
```
```{r}
#Adding perfect intervening at node "SeniorTrain" 
insurance_mut2_network = mutilated(fitted, evidence = list(SeniorTrain = "True"))
jmutilated = compile(as.grain(insurance_mut2_network))

#Sampling experimental data
insurance_mut2_exp8 = rbn(insurance_mut2_network,1000)

#Intitializing additional column for the known intervened target
Int_col = 14

#Adding the INT column
insurance_mut2_exp8$INT = Int_col
```
```{r}
#Combining observation and experiment into one dataframe
insurance_mut2 = rbind(insurance_mut2_obs1, insurance_mut2_obs2, insurance_mut2_obs3, insurance_mut2_exp1, insurance_mut2_exp2, insurance_mut2_exp3, insurance_mut2_exp4, insurance_mut2_exp5, insurance_mut2_exp6, insurance_mut2_exp7, insurance_mut2_exp8)
write.csv(insurance_mut2,"data/insurance data/insurance_mut2.csv",row.names=FALSE)
```
